Implications of afforestation policies for global markets

### 2023-08-07 paper.Rmd
Update paper.Rmd

The global gains that can be realized from afforestation, however, are uncertain, with broad estimates of cumulative sequestration ranging from 42 gigatonnes to 700 gigatonnes by the end of the century [@veldman2019]  [@bastin201]  [@Humpenöder2014]  [@sathaye2006]  [@doelman2020]. In addition, the amount of land conversion that would be required also varies greatly, with estimates ranging from 0.3 billion hectares to 2.8 billion hectares [@veldman2019]  [@bastin201]  [@Humpenöder2014]  [@doelman2020]. In many cases, these results suggest massive losses in the amount of pasture and cropland that would be available. This puts into question the feasibility of attaining these results, as almost 90 percent of deforestation worldwide is undertaken to expand the amount of agricultural land [@fao2021]. There are many reasons for this. For example, as nations develop and disposable incomes rise, consumers often change to more meat-based diets. Population growth is also a driver and when it is especially high, which is the case at present throughout most of Africa, this also puts pressure on policymakers to allow greater conversion of forestland into farmland. Still, previous research has suggested that substantive gains in the carbon sink (+23.8PgCo2e/yr) can be made without sacrificing agricultural land [@griscom2017]. However, as land-use decisions in one country also tend to have impacts that spillover into many others [@meyfroidt2010] [@lambin2011], analysis needs to be conducted at a global scale.

@@

More recent modeling attempts of land-use change have begun to move away from comparative static analysis. For example, the dynamic GTAP model, supported by the GTAP-AEZ database, considers land market effects, which were identified as significant in driving results by Stevenson et al. [@stevenson2013], and has been utilized to evaluate the impacts of crop intensification on land use. It, however, has been critiqued as it global aggregates may mask localized shifts that can have implications for ecologically significant areas [@byerlee2014]. This project will attempt to get around these concerns and assess the impact of various likely afforestation scenarios by melding GTAP with the Global Timber Model (GTM) [@sohngen2001]. The GTM is seen as being quite helpful in getting around these issues, as it is a dynamic optimization model that can examine the impact of afforestation or avoided deforestation on the carbon sink. By joining these two models and passing outputs back and forth between the models, any limitations of either model can be minimized and more robust results, therefore, can be attained.

South America and Sub-Saharan Africa are two regions that show much potential for afforestation. It is estimated that these two regions hold at least 50% of the potential global gain while regions such as Northern Africa and the Middle East show little promise as forest growth rates are quite low [@doelman2020]. Thus, it is promising that the African Forest Landscape Restoration Initiative (AFR100), which aims to afforest 100 million hectares by 2030, has roughly 70% of its commitments coming from Sub-Saharan nations [@auda2023]. While many of these nations may be well-suited in biophysical terms for afforestation, there is concern about how long-lasting any investment may be due to proximate zones of political instability. There are other nations that have a big potential for gains from afforestation, but they also face challenges. India, for example, has a great potential to increase its carbon sequestration through afforestation, but due to its massive population and food security worries, its NDC does not entertain the idea of agricultural land being repurposed and instead primarily considers targeting wastelands that may not be well suited for afforestation [@amjath2019].

>>>>

The global gains that can be realized from afforestation, however, are uncertain, with broad estimates of cumulative sequestration ranging from 42 gigatonnes to 700 gigatonnes by the end of the century [@veldman2019]  [@bastin201]  [@Humpenöder2014]  [@sathaye2006]  [@doelman2020]. In addition, the amount of land conversion that would be required also varies greatly, with estimates ranging from 0.3 billion hectares to 2.8 billion hectares [@veldman2019]  [@bastin201]  [@Humpenöder2014]  [@doelman2020]. In many cases, these results suggest massive losses in the amount of pasture and cropland that would be available. This puts into question the feasibility of attaining these results, as almost 90 percent of deforestation worldwide is undertaken to expand the amount of agricultural land [@fao2021]. There are many reasons for this. For example, as nations develop and disposable incomes rise, consumers often change to more meat-based diets. Population growth is also a driver and when it is especially high, which is the case at present throughout most of Africa, this also puts pressure on policymakers to allow greater conversion of forestland into farmland. Still, previous research has suggested that substantive gains in the carbon sink (+23.8PgCo2e/yr) can be made without sacrificing agricultural land [@griscom2017].

South America and Sub-Saharan Africa are two regions that show much potential for afforestation. It is estimated that these two regions hold at least 50% of the potential global gain while regions such as Northern Africa and the Middle East show little promise as forest growth rates are quite low [@doelman2020]. Thus, it is promising that the African Forest Landscape Restoration Initiative (AFR100), which aims to afforest 100 million hectares by 2030, has roughly 70% of its commitments coming from Sub-Saharan nations [@auda2023]. While many of these nations may be well-suited in biophysical terms for afforestation, there is concern about how long-lasting any investment may be due to proximate zones of political instability. There are other nations that have a big potential for gains from afforestation, but they also face challenges. India, for example, has a great potential to increase its carbon sequestration through afforestation, but due to its massive population and food security worries, its NDC does not entertain the idea of agricultural land being repurposed and instead primarily considers targeting wastelands that may not be well suited for afforestation [@amjath2019]. Moreover, as land-use decisions in one country also tend to have impacts that spillover into many others [@meyfroidt2010] [@lambin2011], analysis needs to be conducted at a global scale.

@@

More recent modeling attempts of land-use change have begun to move away from comparative static analysis. For example, the dynamic GTAP model, supported by the GTAP-AEZ database, considers land market effects, which were identified as significant in driving results by Stevenson et al. [@stevenson2013], and has been utilized to evaluate the impacts of crop intensification on land use. It, however, has been critiqued, as global aggregates may mask localized shifts that can have implications for ecologically significant areas [@byerlee2014]. This project will attempt to get around these concerns and assess the impact of various likely afforestation scenarios by melding GTAP with the Global Timber Model (GTM), which is a dynamic optimization model that can examine the impact of afforestation or avoided deforestation on the carbon sink [@sohngen2001]. By joining these two models and passing outputs back and forth between the models, any limitations of either model can be minimized and more robust results, therefore, can be attained.
